---
title: "Walkthrough of `pls`"
author: "Al Xin"
date: '2022-12-07'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pls)
library(ggplot2)
library(dplyr)
theme_set(theme_bw())
```

## Data loading
```{r}
data(yarn)
data(oliveoil)
data(gasoline)
```

## Gasoline PLS

We first split into training and test sets. 

```{r}
gas_trn <- gasoline[1:50, ]
gas_tst <- gasoline[51:60, ]
```

A standard PLS execution would be
```{r}
gas1 <- plsr(
	octane ~ NIR, 
	ncomp = 10, 
	data = gas_trn, 
	validation = "LOO"
)
```

The following summary statistics includes RMSE of prediction (RMSEP) on the validation set. The adjusted CV is bias-corrected CV; for LOO-CV, it will be very similar to the regular CV value. 

```{r}
summary(gas1)
```

The RMSE plots are below. 

```{r}
library(reshape2)

gas1_rmse <- RMSEP(gas1) %>%
	.$val %>%
	data.frame() %>%
	mutate(cv_type = row.names(.))
colnames(gas1_rmse)[1:11] <- c(0:10)
gas1_rmse <- 	melt(
	gas1_rmse, 
	id.vars = "cv_type",
	value.name = "rmsep", 
	variable.name = "ncomp"
) %>%
	mutate(
		cv_type = as.factor(cv_type), 
		ncomp = as.numeric(ncomp)
	)

ggplot(gas1_rmse, aes(x = ncomp, y = rmsep, color = cv_type)) + 
	geom_line()
```

We can use the bas R `plot` function to visualize predictions from a specified number of components. 

```{r}
plot(gas1, ncomp = 2, asp = 1, line = T)
```

We can also plot the components, giving a pairwose plot of the score values for the first three components. These can be used to look for patterns, groups, or outliers in the data.  

```{r}
plot(gas1, plottype = "scores", comps = 1:3)
```

The loading plot can be used for interpretation. 

```{r}
plot(
	gas1, 
	"loadings", 
	comps = 1:2, 
	legendpos = "topleft", 
	labels = "numbers", 
	xlab = "nm"
)
abline(h = 0)
```

